---
title: "1"
format: html
editor: visual
---

## Call libraries

```{r}
# Load in the packages for this data analysis
librarian::shelf(tidyverse, dplyr, terra, tmap, sf, here, stars, janitor, units, showtext, knitr, kableExtra)
# Adding in our text
font_add_google("Montserrat", "montserrat")
showtext_auto()

options(scipen = 999)
```

## Statewide Analysis:

### Find the total area of state waters using the state water boundary

```{r, message=FALSE}
# Read in the study regions data
state_boundaries <-  sf::st_read("/capstone/marinebiomaps/data/MLPA_Study_Regions")
```

```{r}
# Transform to the UTM 10N crs for California
state_boundary_transform <- st_transform(state_boundaries, crs= 32610)
statewater_area <- sum(st_area(state_boundary_transform) / 10000)
```

```{r}
# Calculate area in hectares
sr_boundaries <- state_boundary_transform |>
  mutate(area_ha = as.numeric(st_area(geometry)) / 10000)
```

```{r}
scsr_area <- 795004.7
ccsr_area <- 296913.5
nccsr_area <- 197503.2
ncsr_area <- 266790.1
```

### Find area in hectares by Substrate Category

```{r}
# Read in the prefiltered sub data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
ca_sub <- readRDS(file.path(rds.dir, "substrate.rds"))


# Remove the unclassified observations
ca_sub_clean <- ca_sub %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop the geometries to speed up run time
ca_sub_simple <- ca_sub_clean %>%
  st_drop_geometry()

# Create a column that has the total area of socal sub
ca_sub_simple <- ca_sub_simple %>%
  mutate(total_area = sum(area_ha) / 10000)
```

```{r}
statewater_area
```

```{r}
unique(ca_sub_simple$total_area)
```

### Get absolute abundance for sub

```{r}
# Create a sum of area column in mpa_transform
ca_sub_simple_2 <- ca_sub_simple %>%
  group_by(cmecs_sc_category) %>%  
  summarise(
    sub_area = sum(area_ha, na.rm = TRUE) / 10000,
  ) %>%
  mutate(pct_sub = (sub_area / statewater_area) * 100) %>% drop_units()
```

```{r}
 # Create table
kable(ca_sub_simple_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total Statewaters")) %>%
  kable_material(c("striped", "hover"))
```

### Repeating our calculations within the MPA system

```{r}
# Load in the MPA boundary data
MPA_boundary <- sf::st_read("/capstone/marinebiomaps/data/MPA_boundaries")

# Filter to Point Lobos locations
mpa <- MPA_boundary %>%
   clean_names() %>% st_make_valid()
```

```{r}
# Reproject to UTM 10
mpa_transform <- st_transform(mpa, crs= 32610)
mpa_area <-  sum(st_area(mpa_transform) / 10000)
```

```{r}
# Join the data sets together
ca_sub_mpa <- st_intersection(ca_sub_clean, mpa_transform)
```

```{r}
ca_sub_mpa$sub_area_ha = (st_area(ca_sub_mpa) / 10000)
```

```{r}
# Remove geometries to run faster
ca_sub_mpa_simple <- ca_sub_mpa %>%
  st_drop_geometry()
```

```{r}
# Create a sum of area column in mpa_transform
ca_sub_mpa_2 <- ca_sub_mpa_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_cat_ha = sum(sub_area_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_sub = (sub_cat_ha / 262882) * 100)
```

```{r}
 # Create table
kable(ca_sub_mpa_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total MPA Area")) %>%
  kable_material(c("striped", "hover"))
```

## Tornado for Statewide

```{r}
# Drop units
ca_sub_simple_2 <- drop_units(ca_sub_simple_2)
ca_sub_mpa_2 <- drop_units(ca_sub_mpa_2)

# Combine dfs to find difference
sub_diff <- full_join(ca_sub_simple_2, ca_sub_mpa_2, by = "cmecs_sc_category") %>%
  select(
    cmecs_sc_category,
    State = pct_sub.x,
    MPA = pct_sub.y
  ) %>%
  mutate(Difference = MPA - State)

# Center chart
max_abs <- max(abs(sub_diff$Difference), na.rm = TRUE)

# Plot
ggplot(sub_diff, aes(x = Difference, y = reorder(str_wrap(cmecs_sc_category, 25), Difference), fill = Difference > 0)) +
  
  # Plot lines and dots
  geom_segment(aes(x = 0, xend = Difference, yend = str_wrap(cmecs_sc_category, 25)), linewidth = 1) +
  geom_point() +
  
  # Create center line at 0
  geom_vline(xintercept = 0, color = "black") +
  
  # Fix x-axis scales
 scale_x_continuous(
  name = "Difference in Substrate (%) (MPA - State)",
  # breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = scales::percent_format(scale = 1, accuracy = 0.1),
  limits = c(-60, 60)) +
  
  # Apply color and labels
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(
    title = "Substrate Composition: MPA vs State Waters",
    y = "") +
  
  # Apply themes
  theme_minimal() +
  
  theme(legend.position = "none",
        plot.title = element_text(size = 20),
        text = element_text(size = 18, family = "montserrat"),
        axis.title.x = element_text(margin = margin(t = 10)))
```

## Abundance in SCSR:

### Region wide:

```{r}
# Make regional boundaries
scsr <- state_boundary_transform %>%
  filter(Study_Regi == "SCSR")

ccsr <- state_boundary_transform %>%
  filter(Study_Regi == "CCSR")

nccsr <- state_boundary_transform %>%
  filter(Study_Regi == "NCCSR")

ncsr <- state_boundary_transform %>%
  filter(Study_Regi == "NCSR")
```

```{r}
# Make valid the scsr geometry
scsr <- scsr %>% 
  st_make_valid()

# Find all sub in scsr region
scsr_sub <- st_intersection(scsr,ca_sub_clean)
```

```{r}
# Remove the unclassified observations
scsr_sub_clean <- scsr_sub %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
scsr_sub_simple <- scsr_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
sub_scsr <- scsr_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(area_ha, na.rm = TRUE) / 10000,
  ) %>%
  mutate(pct_sub = (sub_area /scsr_area) * 100)
```

```{r}
 # Create table
kable(sub_scsr, col.names = c("Substrate Category", "Area in hectares", "Percent of Total SCSR Area")) %>%
  kable_material(c("striped", "hover"))
```

### In SCSR MPAs:

```{r}
# Read in the prefiltered sub data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
socal_sub <- readRDS(file.path(rds.dir, "scsr_substrate.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
socal_sub_transform <- st_transform(socal_sub, crs = 32610)

# Calculate area in hectares
socal_sub_transform$socal_ha <- (st_area(socal_sub_transform) / 10000)

# Remove the unclassified observations
socal_sub_clean <- socal_sub_transform %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
socal_sub_simple <- socal_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
socal_sub_simple_2 <- socal_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(socal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_sub = (sub_area /scsr_area ) * 100)
```

```{r}
 # Create table
kable(socal_sub_simple_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total South Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

## Tornado -SCSR

```{r}
# Drop units
socal_sub_simple_2 <- drop_units(socal_sub_simple_2)
sub_scsr <- drop_units(sub_scsr)

# Combine dfs to find difference
sub_diff2 <- full_join(socal_sub_simple_2, sub_scsr, by = "cmecs_sc_category") %>%
  select(
    cmecs_sc_category,
    State = pct_sub.x,
    MPA = pct_sub.y
  ) %>%
  mutate(Difference = MPA - State)

# Center chart
max_abs <- max(abs(sub_diff$Difference), na.rm = TRUE)

# Plot
ggplot(sub_diff2, aes(x = Difference, y = reorder(str_wrap(cmecs_sc_category, 25), Difference), fill = Difference > 0)) +
  
  # Plot lines and dots
  geom_segment(aes(x = 0, xend = Difference, yend = str_wrap(cmecs_sc_category, 25)), size = 1) +
  geom_point() +
  
  # Create center line at 0
  geom_vline(xintercept = 0, color = "black") +
  
  # Fix x-axis scales
 scale_x_continuous(
  name = "Difference in Substrate (%) (MPA - State)",
#  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = scales::percent_format(scale = 1, accuracy = 0.1),
  limits = c(-11, 11)) +
  
  # Apply color and labels
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(
    title = "Substrate Composition: SCSR MPA vs State Waters",
    y = "") +
  
  # Apply themes
  theme_minimal() +
  
  theme(legend.position = "none",
        plot.title = element_text(size = 20),
        text = element_text(size = 18, family = "montserrat"),
        axis.title.x = element_text(margin = margin(t = 10)))
```

## Abundance in CCSR:

### In CCSR region

```{r}
# Make valid the scsr geometry
ccsr <- ccsr %>% 
  st_make_valid()

# Find all sub in scsr region
ccsr_sub <- st_intersection(ccsr,ca_sub_clean)
```

```{r}
# Remove the unclassified observations
ccsr_sub_clean <- ccsr_sub %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
ccsr_sub_simple <- ccsr_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
sub_ccsr <- ccsr_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(area_ha, na.rm = TRUE) / 10000,
  ) %>%
  mutate(pct_sub = (sub_area /ccsr_area) * 100)
```

```{r}
 # Create table
kable(sub_ccsr, col.names = c("Substrate Category", "Area in hectares", "Percent of Total Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In CCSR MPAs:

```{r}
# Read in the prefiltered sub data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
central_sub <- readRDS(file.path(rds.dir, "ccsr_substrate.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
cencal_sub_transform <- st_transform(central_sub, crs = 32610)

# Calculate area in hectares
cencal_sub_transform$cencal_ha <- (st_area(cencal_sub_transform) / 10000)

# Remove the unclassified observations
cencal_sub_clean <- cencal_sub_transform %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
cencal_sub_simple <- cencal_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
cencal_sub_simple_2 <- cencal_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(cencal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_sub = (sub_area /ccsr_area ) * 100)
```

```{r}
 # Create table
kable(cencal_sub_simple_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total Central Coast Region MPAS")) %>%
  kable_material(c("striped", "hover"))
```

## Tornado CCSR

```{r}
# Drop units
cencal_sub_simple_2 <- drop_units(cencal_sub_simple_2)
sub_ccsr <- drop_units(sub_ccsr)

# Combine dfs to find difference
sub_diff3 <- full_join(cencal_sub_simple_2, sub_ccsr, by = "cmecs_sc_category") %>%
  select(
    cmecs_sc_category,
    State = pct_sub.x,
    MPA = pct_sub.y) %>%
    mutate(
    MPA = replace_na(MPA, 0),            
    State = replace_na(State, 0),        
    Difference = MPA - State)

# Center chart
max_abs <- max(abs(sub_diff$Difference), na.rm = TRUE)

# Plot
ggplot(sub_diff3, aes(x = Difference, y = reorder(str_wrap(cmecs_sc_category, 25), Difference), fill = Difference > 0)) +
  
  # Plot lines and dots
  geom_segment(aes(x = 0, xend = Difference, yend = str_wrap(cmecs_sc_category, 25)), linewidth = 1) +
  geom_point() +
  
  # Create center line at 0
  geom_vline(xintercept = 0, color = "black") +
  
  # Fix x-axis scales
 scale_x_continuous(
  name = "Difference in Substrate (%) (MPA - State)",
#  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = scales::percent_format(scale = 1, accuracy = 0.1),
  limits = c(-11, 11)) +
  
  # Apply color and labels
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(
    title = "Substrate Composition: CCSR MPA vs State Waters",
    y = "") +
  
  # Apply themes
  theme_minimal() +
  
  theme(legend.position = "none",
        plot.title = element_text(size = 20),
        text = element_text(size = 18, family = "montserrat"),
        axis.title.x = element_text(margin = margin(t = 10)))
```


## Abundance in NCCSR:

### In NCCSR Region

```{r}
# Make valid the scsr geometry
nccsr <- nccsr %>% 
  st_make_valid()

# Find all sub in scsr region
nccsr_sub <- st_intersection(nccsr,ca_sub_clean)
```

```{r}
# Remove the unclassified observations
nccsr_sub_clean <- nccsr_sub %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
nccsr_sub_simple <- nccsr_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
sub_nccsr <- nccsr_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(area_ha, na.rm = TRUE) / 10000,
  ) %>%
  mutate(pct_sub = (sub_area /nccsr_area) * 100)
```

```{r}
 # Create table
kable(sub_nccsr, col.names = c("Substrate Category", "Area in hectares", "Percent of Total North Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In NCCSR MPAs

```{r}
# Read in the prefiltered sub data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
nccsr_sub <- readRDS(file.path(rds.dir, "nccsr_substrate.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
nccsr_sub_transform <- st_transform(nccsr_sub, crs = 32610)

# Calculate area in hectares
nccsr_sub_transform$nccsr_ha <- (st_area(nccsr_sub_transform) / 10000)

# Remove the unclassified observations
nccsr_sub_clean <- nccsr_sub_transform %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
nccsr_sub_simple <- nccsr_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column 
nccsr_sub_simple_2 <- nccsr_sub_simple %>%
  group_by(cmecs_sc_category) %>%  
  summarise(
    sub_area = sum(nccsr_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_sub = (sub_area /nccsr_area ) * 100)
```

```{r}
 # Create table
kable(nccsr_sub_simple_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total North Central Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

## Tornado NCSR

```{r}
# Drop units
nccsr_sub_simple_2 <- drop_units(nccsr_sub_simple_2)
sub_nccsr <- drop_units(sub_nccsr)

# Combine dfs to find difference
sub_diff4 <- full_join(nccsr_sub_simple_2, sub_nccsr, by = "cmecs_sc_category") %>%
  select(
    cmecs_sc_category,
    State = pct_sub.x,
    MPA = pct_sub.y
  ) %>%
  mutate(Difference = MPA - State)

# Center chart
max_abs <- max(abs(sub_diff$Difference), na.rm = TRUE)

# Plot
ggplot(sub_diff4, aes(x = Difference, y = reorder(str_wrap(cmecs_sc_category, 25), Difference), fill = Difference > 0)) +
  
  # Plot lines and dots
  geom_segment(aes(x = 0, xend = Difference, yend = str_wrap(cmecs_sc_category, 25)), linewidth = 1) +
  geom_point() +
  
  # Create center line at 0
  geom_vline(xintercept = 0, color = "black") +
  
  # Fix x-axis scales
 scale_x_continuous(
  name = "Difference in Substrate (%) (MPA - State)",
#  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = scales::percent_format(scale = 1, accuracy = 0.1),
  limits = c(-8, 8)) +
  
  # Apply color and labels
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(
    title = "Substrate Composition: NCCSR MPA vs State Waters",
    y = "") +
  
  # Apply themes
  theme_minimal() +
  
  theme(legend.position = "none",
        plot.title = element_text(size = 20),
        text = element_text(size = 18, family = "montserrat"),
        axis.title.x = element_text(margin = margin(t = 10)))
```

## Abundance in NCSR:

### In the NCSR Region:

```{r}
# Make valid the scsr geometry
ncsr <- ncsr %>% 
  st_make_valid()

# Find all sub in scsr region
ncsr_sub <- st_intersection(ncsr,ca_sub_clean)
```

```{r}
# Remove the unclassified observations
ncsr_sub_clean <- ncsr_sub %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
ncsr_sub_simple <- ncsr_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
sub_ncsr <- ncsr_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(area_ha, na.rm = TRUE) / 10000,
  ) %>%
  mutate(pct_sub = (sub_area /ncsr_area) * 100)
```

```{r}
 # Create table
kable(sub_ncsr, col.names = c("Substrate Category", "Area in hectares", "Percent of Total North Coast Region")) %>%
  kable_material(c("striped", "hover"))
```

### In NCSR MPAs

```{r}
# Read in the prefiltered sub data from the database
rds.dir <- "/capstone/marinebiomaps/data/rds-files"
ncsr_sub <- readRDS(file.path(rds.dir, "ncsr_substrate.rds"))
```

```{r}
# Reproject to crs for WGS84 UTM zone 10N
norcal_sub_transform <- st_transform(ncsr_sub, crs = 32610)

# Calculate area in hectares
norcal_sub_transform$norcal_ha <- (st_area(norcal_sub_transform) / 10000)

# Remove the unclassified observations
norcal_sub_clean <- norcal_sub_transform %>%
  filter(cmecs_sc_category != "Unclassified")

# Drop geometries for ease of processing
norcal_sub_simple <- norcal_sub_clean %>%
  st_drop_geometry()

# Create a sum of area column in mpa_transform
norcal_sub_simple_2 <- norcal_sub_simple %>%
  group_by(cmecs_sc_category) %>%  # Group by MPA type, substrate group
  summarise(
    sub_area = sum(norcal_ha, na.rm = TRUE),
  ) %>%
  mutate(pct_sub = (sub_area /ncsr_area ) * 100)
```

```{r}
 # Create table
kable(norcal_sub_simple_2, col.names = c("Substrate Category", "Area in hectares", "Percent of Total North Coast Region MPAs")) %>%
  kable_material(c("striped", "hover"))
```

## Tornado NCSR

```{r}
# Drop units
norcal_sub_simple_2 <- drop_units(norcal_sub_simple_2)
sub_ncsr <- drop_units(sub_ncsr)

# Combine dfs to find difference
sub_diff5 <- full_join(norcal_sub_simple_2, sub_ncsr, by = "cmecs_sc_category") %>%
  select(
    cmecs_sc_category,
    State = pct_sub.x,
    MPA = pct_sub.y) %>%
    mutate(
    MPA = replace_na(MPA, 0),            
    State = replace_na(State, 0),        
    Difference = MPA - State)

# Center chart
max_abs <- max(abs(sub_diff$Difference), na.rm = TRUE)

# Plot
ggplot(sub_diff5, aes(x = Difference, y = reorder(str_wrap(cmecs_sc_category, 25), Difference), fill = Difference > 0)) +
  
  # Plot lines and dots
  geom_segment(aes(x = 0, xend = Difference, yend = str_wrap(cmecs_sc_category, 25)), linewidth = 1) +
  geom_point() +
  
  # Create center line at 0
  geom_vline(xintercept = 0, color = "black") +
  
  # Fix x-axis scales
 scale_x_continuous(
  name = "Difference in Substrate (%) (MPA - State)",
#  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = scales::percent_format(scale = 1, accuracy = 0.1),
  limits = c(-7, 7)) +
  
  # Apply color and labels
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(
    title = "Substrate Composition: NCSR MPA vs State Waters",
    y = "") +
  
  # Apply themes
  theme_minimal() +
  
  theme(legend.position = "none",
        plot.title = element_text(size = 20),
        text = element_text(size = 18, family = "montserrat"),
        axis.title.x = element_text(margin = margin(t = 10)))
```
